Good Empirics Dominate Bad Theory: Comments on Blair and Sambanis, 2020, ``Forecasting Civil Wars: Theory and Structure in an Age of `Big Data' and Machine Learning'', \emph{JCR}

### 2020-06-03 paper.Rmd
wednesday mdw
some changes

library(readr)
library(dplyr)
library(tidyr)
library(tidyverse)
library(kableExtra)
library(stringr)
```

@@

- (1) how does the theoretically-driven Escalation model compare in forecast performance to alternative models not informed specifically by civil war onset theories, 
- (2) does annual, structural information from the PITF instability forecasting model add to the Escalation model's monthly and 6-month predictions,
- and (3) how accurate were predictions for the first half of 2016 derived from the Escalation model?

@@

To examine the first question, BS compare the test set of the Escalation model to four alternative models. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The models are:

- Escalation: a set of 10 theoretically-informed indicators.
- Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government one one side and opposition or rebel actors on the other. These are directed, thus making for 4 total covariates.
- CAMEO: counts for all CAMEO event codes, thus a total of 1,159 covariates. 
- Average: unweighted average of the predictions from the 4 models above.

>>>>

library(readr)
library(dplyr)
library(tidyr)
library(tidyverse)
library(kableExtra)
library(stringr)
```
# Introduction

@bdm:2011 , @ward:2016 , @montgomery:etal:PS:2012, @chiba:etal:2015, @brandt:2015  @mhw:2012,@ward:etal:2013,@chadefaux:2015,@weidmann:ward:2010,@pilster:bohmelt:2014,@Brandt_etal_2011,@gleditsch:2015,@bennett:stam:2009,@shellman:etal:2013,@hegre:etal:2011,@ward:etal:2010,@freeman:job:1979,@gneiting:raftery:2005
@choucri:1974,@choucri:robinson:1979,@goldstone:etal:2010,tetlock:etal:2017,

@@

- (1) How does the theoretically-driven rscalation model compare in forecast performance to alternative models not informed specifically by civil war onset theories?
- (2) Does annual, structural information from the PITF instability forecasting model add to the escalation model's monthly and 6-month predictions?
- (3) How accurate were predictions for the first half of 2016 derived from the escalation model?

@@

To examine the first question, BS compare the test set of the escalation model to four alternative models. The independent variables for the first set of analysis reported in Table 1 in the paper are all derived from the ICEWS event data, using domestic events between actors within a country. The models are:

- Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model.
- Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation.
- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for interactions between the government one one side and opposition or rebel actors on the other. These are directed, thus making for 4 total covariates.
- CAMEO: counts for all CAMEO event codes, thus a total of 1,159 covariates (CAN THIS BE RIGHT?). 
- Average: unweighted average of the predictions from the four models briefly described above.

### 2020-06-03 paper.tex
wednesday mdw
some changes

@@

\begin{itemize}
\item
  \begin{enumerate}
  \def\labelenumi{(\arabic{enumi})}
  \tightlist
  \item
    how does the theoretically-driven Escalation model compare in
    forecast performance to alternative models not informed specifically
    by civil war onset theories,
  \end{enumerate}
\item
  \begin{enumerate}
  \def\labelenumi{(\arabic{enumi})}
  \setcounter{enumi}{1}
  \tightlist
  \item
    does annual, structural information from the PITF instability
    forecasting model add to the Escalation model's monthly and 6-month
    predictions,
  \end{enumerate}
\item
  and (3) how accurate were predictions for the first half of 2016
  derived from the Escalation model?
\end{itemize}

@@

To examine the first question, BS compare the test set of the Escalation
model to four alternative models. The independent variables for the
first set of analysis reported in Table 1 in the paper are all derived
from the ICEWS event data, using domestic events between actors within a
country. The models are:

\begin{itemize}
\tightlist
\item
  Escalation: a set of 10 theoretically-informed indicators.
\item
  Quad: ICEWS quad counts, i.e.~material conflict, material cooperation,
  verbal conflict, verbal cooperation.
\item
  Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from
  the ICEWS data for interactions between the government one one side
  and opposition or rebel actors on the other. These are directed, thus
  making for 4 total covariates.
\item
  CAMEO: counts for all CAMEO event codes, thus a total of 1,159
  covariates.
\item
  Average: unweighted average of the predictions from the 4 models
  above.
\end{itemize}

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-blair2020forecasting}{}%
Blair, Robert A., and Nicholas Sambanis. 2020. ``Forecasting Civil Wars:
Theory and Structure in an Age of `Big Data' and Machine Learning.''
\emph{Journal of Conflict Resolution}.

>>>>

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Bueno de Mesquita (2011) , Ward (2016) , Montgomery, Hollenbach, and
Ward (2012a), Chiba, Metternich, and Ward (2015), Brandt (2015)
Montgomery, Hollenbach, and Ward (2012b),Ward et al. (2013),Chadefaux
(2015),Weidmann and Ward (2010),Pilster and Böhmelt (2014),Brandt,
Freeman, and Schrodt (2011),Gleditsch (2015),Bennett and Stam
(2009),Shellman, Levey, and Young (2013),Hegre, Hultman, and Nygård
(2011),Ward, Greenhill, and Bakke (2010),Freeman and Job (1979),Gneiting
and Raftery (2005) Choucri (1974),Choucri and Robinson (1979),Goldstone
et al. (2010),tetlock:etal:2017,

@@

\begin{itemize}
\item
  \begin{enumerate}
  \def\labelenumi{(\arabic{enumi})}
  \tightlist
  \item
    How does the theoretically-driven rscalation model compare in
    forecast performance to alternative models not informed specifically
    by civil war onset theories?
  \end{enumerate}
\item
  \begin{enumerate}
  \def\labelenumi{(\arabic{enumi})}
  \setcounter{enumi}{1}
  \tightlist
  \item
    Does annual, structural information from the PITF instability
    forecasting model add to the escalation model's monthly and 6-month
    predictions?
  \end{enumerate}
\item
  \begin{enumerate}
  \def\labelenumi{(\arabic{enumi})}
  \setcounter{enumi}{2}
  \tightlist
  \item
    How accurate were predictions for the first half of 2016 derived
    from the escalation model?
  \end{enumerate}
\end{itemize}

@@

To examine the first question, BS compare the test set of the escalation
model to four alternative models. The independent variables for the
first set of analysis reported in Table 1 in the paper are all derived
from the ICEWS event data, using domestic events between actors within a
country. The models are:

\begin{itemize}
\tightlist
\item
  Escalation: a set of ten indicators, putatively drawn from a
  theoretical escalation model.
\item
  Quad: ICEWS quad counts, i.e.~material conflict, material cooperation,
  verbal conflict, verbal cooperation.
\item
  Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from
  the ICEWS data for interactions between the government one one side
  and opposition or rebel actors on the other. These are directed, thus
  making for 4 total covariates.
\item
  CAMEO: counts for all CAMEO event codes, thus a total of 1,159
  covariates (CAN THIS BE RIGHT?).
\item
  Average: unweighted average of the predictions from the four models
  briefly described above.
\end{itemize}

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-bennett:stam:2009}{}%
Bennett, D. Scott, and Allan C. Stam. 2009. ``Revisiting Predictions of
War Duration.'' \emph{Conflict Management and Peace Science} 26 (3):
256--67.

\leavevmode\hypertarget{ref-blair2020forecasting}{}%
Blair, Robert A., and Nicholas Sambanis. 2020. ``Forecasting Civil Wars:
Theory and Structure in an Age of `Big Data' and Machine Learning.''
\emph{Journal of Conflict Resolution}.

\leavevmode\hypertarget{ref-brandt:2015}{}%
Brandt, Patrick. 2015. ``Forecasting Conflicts: Long and Short Term
Predictions Based on Different Training Set Considerations.'' University
of Texas Dallas; Peace Research Institute Oslo.

\leavevmode\hypertarget{ref-Brandt_etal_2011}{}%
Brandt, Patrick T., John R. Freeman, and Philip A. Schrodt. 2011. ``Real
Time, Time Series Forecasting of Inter- and Intra-State Political
Conflict.'' \emph{Conflict Management and Peace Science} 28 (1): 41--64.

\leavevmode\hypertarget{ref-bdm:2011}{}%
Bueno de Mesquita, Bruce. 2011. ``A New Model for Predicting Policy
Choices: Preliminary Tests.'' \emph{Conflict Management and Peace
Science} 28 (1): 65--85.

\leavevmode\hypertarget{ref-chadefaux:2015}{}%
Chadefaux, Thomas. 2015. ``Predictably Unpredictable: The Limits of
Conflict Forecasting.''

\leavevmode\hypertarget{ref-chiba:etal:2015}{}%
Chiba, Daina, Nils W. Metternich, and Michael D. Ward. 2015. ``Every
Story Has a Beginning, Middle, and an End (but Not Always in That
Order): Predicting Duration Dynamics in a Unified Framework.''
\emph{Political Science Research and Methods} 3 (3): 515--41.

\leavevmode\hypertarget{ref-choucri:1974}{}%
Choucri, Nazli. 1974. ``Forecasting in International Relations: Problems
and Prospects.'' \emph{International Interactions} 1 (2): 63--68.

\leavevmode\hypertarget{ref-choucri:robinson:1979}{}%
Choucri, Nazli, and Thomas W. Robinson, eds. 1979. \emph{Forecasting in
International Relations: Theory, Methods, Problems, and Prospects}. San
Francisco, CA: W.H. Freeman.

\leavevmode\hypertarget{ref-freeman:job:1979}{}%
Freeman, John R., and Brian L. Job. 1979. ``Scientific Forecasts in
International Relations: Problems of Definition and Epistemology.''
\emph{International Studies Quarterly} 23 (1). Blackwell Publishing on
behalf of The International Studies Association: 113--43.

\leavevmode\hypertarget{ref-gleditsch:2015}{}%
Gleditsch, Kristian Skrede. 2015. ``Predicting Ucdp/Prio Civil Wars:
Expanding the Inequality and Grievance Model with Event Data.''
University of Essex; Peace Research Institute Oslo.

\leavevmode\hypertarget{ref-gneiting:raftery:2005}{}%
Gneiting, Tilman, and Adrian E. Raftery. 2005. ``Weather Forecasting
with Ensemble Methods.'' \emph{Science} 310: 248--49.

\leavevmode\hypertarget{ref-goldstone:etal:2010}{}%
Goldstone, Jack A, Robert H Bates, David L Epstein, Ted Robert Gurr,
Michael B Lustik, Monty G Marshall, Jay Ulfelder, and Mark Woodward.
2010. ``A Global Model for Forecasting Political Instability.''
\emph{American Journal of Political Science} 54 (1). Wiley Online
Library: 190--208.

\leavevmode\hypertarget{ref-hegre:etal:2011}{}%
Hegre, Håvard, Lisa Hultman, and Håvard Mokleiv Nygård. 2011.
``Simulating the Effect of Peacekeeping Operations 2010--2035.'' In
\emph{Social Computing, Behavioral-Cultural Modeling and Prediction},
edited by John Salerno, Shanchieh Jay Yang, Dana Nau, and Sun-Ki Chai.,
325--32. Lecture Notes in Computer Science 6589. Springer.

\leavevmode\hypertarget{ref-montgomery:etal:PS:2012}{}%
Montgomery, Jacob M., Florian M. Hollenbach, and Michael D. Ward. 2012a.
``Ensemble Predictions of the 2012 Us Presidential Election.'' \emph{PS:
Political Science \& Politics} 45: 651--54.

\leavevmode\hypertarget{ref-mhw:2012}{}%
Montgomery, Jacob M., Florian Hollenbach, and Michael D. Ward. 2012b.
``Improving Predictions Using Ensemble Bayesian Model Averaging.''
\emph{Political Analysis} 20 (3): 271--91.

\leavevmode\hypertarget{ref-pilster:bohmelt:2014}{}%
Pilster, Ulrich, and Tobias Böhmelt. 2014. ``Predicting the Duration of
the Syrian Insurgency.'' \emph{Research \& Politics} 1 (2).

\leavevmode\hypertarget{ref-shellman:etal:2013}{}%
Shellman, Stephen M., Brian P. Levey, and Joseph K. Young. 2013.
``Shifting Sands: Explaining and Predicting Phase Shifts by Dissident
Organizations.'' \emph{Journal of Peace Research} 50 (3). SAGE
Publications: 319--36.

\leavevmode\hypertarget{ref-ward:2016}{}%
Ward, Michael D. 2016. ``Can We Predict Politics? Toward What End?''
\emph{Journal of Global Security Studies} 1 (1): 80--91.

\leavevmode\hypertarget{ref-ward:etal:2010}{}%
Ward, Michael D., Brian D. Greenhill, and Kristin M. Bakke. 2010. ``The
Perils of Policy by P-Value: Predicting Civil Conflicts.'' \emph{Journal
of Peace Research} 47 (4): 363--75.

\leavevmode\hypertarget{ref-ward:etal:2013}{}%
Ward, Michael D., Nils W. Metternich, Cassy L. Dorff, Max Gallop,
Florian M. Hollenbach, Anna Schultz, and Simon Weschle. 2013. ``Learning
from the Past and Stepping into the Future: Toward a New Generation of
Conflict Prediction.'' \emph{International Studies Review} 16 (4):
473--644.

\leavevmode\hypertarget{ref-weidmann:ward:2010}{}%
Weidmann, N.B., and M.D. Ward. 2010. ``Predicting Conflict in Space and
Time.'' \emph{Journal of Conflict Resolution} 54 (6). SAGE Publications:
883--901.
